Titanic analysis
========================================================

First, create some helper functions

```{r}
readData <- function(path.name, file.name, column.types, missing.types) {
  read.csv( paste(path.name, file.name, sep=""), 
            colClasses=column.types,
            na.strings=missing.types )
}

getTitle <- function(data) {
  title.dot.start <- regexpr("\\,[A-Z ]{1,20}\\.", data$Name, TRUE)
  title.comma.end <- title.dot.start + attr(title.dot.start, "match.length")-1
  data$Title <- substr(data$Name, title.dot.start+2, title.comma.end-1)
  return (data$Title)
}

Titanic.path <- "~/Developer/Kaggle/Titanic/"
train.data.file <- "train.csv"
test.data.file <- "test.csv"
missing.types <- c("NA", "")
train.column.types <- c('integer',   # PassengerId
                        'factor',    # Survived
                        'factor',    # Pclass
                        'character', # Name
                        'factor',    # Sex
                        'numeric',   # Age
                        'integer',   # SibSp
                        'integer',   # Parch
                        'character', # Ticket
                        'numeric',   # Fare
                        'character', # Cabin
                        'factor'     # Embarked
)
test.column.types <- train.column.types[-2]
train.raw <- readData(Titanic.path, train.data.file,
                      train.column.types, missing.types)
df.train <- train.raw
test.raw <- readData(Titanic.path, test.data.file,
                     test.column.types, missing.types)
df.infer <- test.raw

require(Amelia)
missmap(df.train, main="Titanic Training Data - Missings Map",
        col = c("yellow", "black"), legend = FALSE)

df.train$Title <- getTitle(df.train)
```

Consolidating the various titles

```{r}
changeTitles <- function(data, old.titles, new.title) {
  for (honorific in old.titles) {
    data$New.Title[ which( data$Title == honorific)] <- new.title
  }
  return (data$New.Title)
}
df.train$New.Title <- df.train$Title
df.train$New.Title <- changeTitles(df.train, 
                               c("Capt", "Col", "Don", "Dr", 
                               "Jonkheer", "Lady", "Major", 
                               "Rev", "Sir"),
                               "Noble")
df.train$New.Title <- changeTitles(df.train, c("the Countess", "Ms"), 
                               "Mrs")
df.train$New.Title <- changeTitles(df.train, c("Mlle", "Mme"), "Miss")
df.train$New.Title <- as.factor(df.train$New.Title)
```

I'm going to try to get average age based on NEW title and class

```{r}
df.train$TitleClass <- paste0(df.train$New.Title, df.train$Pclass)
avg.age.by.titleclass <- tapply(df.train$Age, df.train$TitleClass, FUN=mean, na.rm=TRUE)
df.train$Impute.Age <- avg.age.by.titleclass[df.train$TitleClass]
df.train$New.Age <- df.train$Age
df.train$New.Age[which(is.na(df.train$Age))] <-
  df.train$Impute.Age[which(is.na(df.train$Age))]

df.train$Fare.Temp <- df.train$Fare
df.train$Fare.Temp[which(df.train$Fare == 0)] <- NA
avg.fare.by.titleclass <- tapply(df.train$Fare.Temp, df.train$TitleClass, FUN=mean, na.rm=TRUE)
df.train$Impute.Fare <- avg.fare.by.titleclass[df.train$TitleClass]
df.train$New.Fare <- df.train$Fare
df.train$New.Fare[which(df.train$Fare == 0)] <-
  df.train$Impute.Fare[which(df.train$Fare == 0)]
```

Then a whole bunch more feature re-engineering (not all of which I necessarily agree with at this point...)

```{r}
require(plyr)     # for the revalue function 
require(stringr)  # for the str_sub function

## test a character as an EVEN single digit
isEven <- function(x) x %in% c("0","2","4","6","8") 
## test a character as an ODD single digit
isOdd <- function(x) x %in% c("1","3","5","7","9") 

## function to add features to training or test data frames
featureEngrg <- function(data) {
  ## Using Fate ILO Survived because term is shorter and just sounds good
  data$Fate <- data$Survived
  ## Revaluing Fate factor to ease assessment of confusion matrices later
  data$Fate <- revalue(data$Fate, c("1" = "Survived", "0" = "Perished"))
  ## Boat.dibs attempts to capture the "women and children first"
  ## policy in one feature.  Assuming all females plus males under 15
  ## got "dibs' on access to a lifeboat
  data$Boat.Dibs <- "No"
  data$Boat.Dibs[which(data$Sex == "female" | data$Age < 15)] <- "Yes"
  data$Boat.Dibs <- as.factor(data$Boat.Dibs)
  ## Family consolidates siblings and spouses (SibSp) plus
  ## parents and children (Parch) into one feature
  data$Family <- data$SibSp + data$Parch
  ## Fare.pp attempts to adjust group purchases by size of family
  data$Fare.pp <- data$New.Fare/(data$Family + 1)
  ## Giving the traveling class feature a new look
  data$Class <- data$Pclass
  data$Class <- revalue(data$Class, 
                        c("1"="First", "2"="Second", "3"="Third"))
  ## First character in Cabin number represents the Deck 
  data$Deck <- substring(data$Cabin, 1, 1)
  data$Deck[ which( is.na(data$Deck ))] <- "UNK"
  data$Deck <- as.factor(data$Deck)
  ## Odd-numbered cabins were reportedly on the port side of the ship
  ## Even-numbered cabins assigned Side="starboard"
  data$Cabin.Last.Digit <- str_sub(data$Cabin, -1)
  data$Side <- "UNK"
  data$Side[which(isEven(data$Cabin.Last.Digit))] <- "port"
  data$Side[which(isOdd(data$Cabin.Last.Digit))] <- "starboard"
  data$Side <- as.factor(data$Side)
  data$Cabin.Last.Digit <- NULL
  return (data)
}

## function to add features to training or test data frames
featureEngrg2 <- function(data) {
  ## Boat.dibs attempts to capture the "women and children first"
  ## policy in one feature.  Assuming all females plus males under 15
  ## got "dibs' on access to a lifeboat
  data$Boat.Dibs <- "No"
  data$Boat.Dibs[which(data$Sex == "female" | data$Age < 15)] <- "Yes"
  data$Boat.Dibs <- as.factor(data$Boat.Dibs)
  ## Family consolidates siblings and spouses (SibSp) plus
  ## parents and children (Parch) into one feature
  data$Family <- data$SibSp + data$Parch
  ## Fare.pp attempts to adjust group purchases by size of family
  data$Fare.pp <- data$New.Fare/(data$Family + 1)
  ## Giving the traveling class feature a new look
  data$Class <- data$Pclass
  data$Class <- revalue(data$Class, 
                        c("1"="First", "2"="Second", "3"="Third"))
  ## First character in Cabin number represents the Deck 
  data$Deck <- substring(data$Cabin, 1, 1)
  data$Deck[ which( is.na(data$Deck ))] <- "UNK"
  data$Deck <- as.factor(data$Deck)
  ## Odd-numbered cabins were reportedly on the port side of the ship
  ## Even-numbered cabins assigned Side="starboard"
  data$Cabin.Last.Digit <- str_sub(data$Cabin, -1)
  data$Side <- "UNK"
  data$Side[which(isEven(data$Cabin.Last.Digit))] <- "port"
  data$Side[which(isOdd(data$Cabin.Last.Digit))] <- "starboard"
  data$Side <- as.factor(data$Side)
  data$Cabin.Last.Digit <- NULL
  return (data)
}

## not sure I love this, but OK
df.train$Embarked[which(is.na(df.train$Embarked))] <- 'S'

## add remaining features to training data frame
df.train <- featureEngrg(df.train)
train.keeps <- c("Fate", "Sex", "Boat.Dibs", "New.Age", "New.Title", "Class", "Deck", "Side", "New.Fare", "Fare.pp", "Embarked", "Family")
df.train.munged <- df.train[train.keeps]
```

Further split the training set into test / training

```{r}
require(caret)
set.seed(23)
training.rows <- createDataPartition(df.train.munged$Fate, 
                                     p = 0.8, list = FALSE)
train.batch <- df.train.munged[training.rows, ]
test.batch <- df.train.munged[-training.rows, ]
```

```{r}
titanic.logit.1 <- glm(Fate ~ Sex + Class + New.Age + Family + Embarked + New.Fare, data = train.batch, family = binomial("logit"))
anova(titanic.logit.1, test="Chisq")
```

```{r}
titanic.logit.2 <- glm(Fate ~ Sex + Class + New.Title + Family  + New.Age+ Embarked, data = train.batch, family = binomial("logit"))
anova(titanic.logit.2, test="Chisq")
```

Apparently the following allows you to get a sensitivity curve for various threshhold values

```{r}
require(pROC)
cv.ctrl <- trainControl(method = "repeatedcv", repeats = 3,
                        summaryFunction = twoClassSummary,
                        classProbs = TRUE)
set.seed(35)
glm.tune.1 <- train(Fate ~ Sex + Class + New.Age + Family + Embarked,
                    data = train.batch,
                    method = "glm",
                    metric = "ROC",
                    trControl = cv.ctrl)
glm.tune.2 <- train(Fate ~ Sex + Class + Sex * Class + New.Age + Family + Embarked,
                    data = train.batch,
                    method = "glm",
                    metric = "ROC",
                    trControl = cv.ctrl)
glm.tune.3 <- train(Fate ~ Sex + Class + Sex * Class + New.Title + New.Age + Family + Embarked,
                    data = train.batch,
                    method = "glm",
                    metric = "ROC",
                    trControl = cv.ctrl)
glm.tune.4 <- train(Fate ~ Sex + Class + Sex * Class + New.Title + New.Age + Family,
                    data = train.batch,
                    method = "glm",
                    metric = "ROC",
                    trControl = cv.ctrl)
glm.tune.5 <- train(Fate ~ Class + I(New.Title=="Mr") + I(New.Title=="Noble") + New.Age + Family + I(Embarked=="S") + I(New.Title=="Mr"&Class=="Third"),
                    data = train.batch,
                    method = "glm",
                    metric = "ROC",
                    trControl = cv.ctrl)
rf.grid <- data.frame(.mtry = c(2, 3))
set.seed(35)
rf.tune <- train(Fate ~ Sex + Class + New.Title + New.Age + Family + Embarked, 
                 data = train.batch,
                 method = "rf",
                 metric = "ROC",
                 tuneGrid = rf.grid,
                 trControl = cv.ctrl)
```

OK, let's try to munge the test file

```{r}
# get titles
df.infer$Title <- getTitle(df.infer)

# I'll consolidate the titles before getting using them to get age
df.infer$New.Title <- df.infer$Title
df.infer$New.Title <- changeTitles(df.infer, c("Dona", "Ms"), "Mrs")
df.infer$New.Title <- changeTitles(df.infer, c("Col", "Dr", "Rev"), "Noble")
df.infer$New.Title <- changeTitles(df.infer, c("Mlle", "Mme"), "Miss")
df.infer$New.Title <- as.factor(df.infer$New.Title)

# Create the title-class variable and use to impute fare and age
df.infer$TitleClass <- paste0(df.infer$New.Title, df.infer$Pclass)

mean.age.by.titleclass <- tapply(df.infer$Age, df.infer$TitleClass, mean, na.rm=TRUE)
df.infer$Impute.Age <- mean.age.by.titleclass[df.infer$TitleClass]
df.infer$New.Age <- df.infer$Age
df.infer$New.Age[which(is.na(df.infer$Age))] <-
  df.infer$Impute.Age[which(is.na(df.infer$Age))]

# Now impute fare
df.infer$Fare.Temp <- df.infer$Fare
df.infer$Fare.Temp[which(df.infer$Fare == 0)] <- NA
avg.fare.by.titleclass <- tapply(df.infer$Fare.Temp, df.infer$TitleClass, FUN=mean, na.rm=TRUE)
df.infer$Impute.Fare <- avg.fare.by.titleclass[df.infer$TitleClass]
df.infer$New.Fare <- df.infer$Fare
df.infer$New.Fare[which(df.infer$Fare == 0)] <-
  df.infer$Impute.Fare[which(df.infer$Fare == 0)]

df.infer <- featureEngrg2(df.infer)

test.keeps <- train.keeps[-1]
pred.these <- df.infer[test.keeps]

g

Survived <- predict(glm.tune.5, newdata = pred.these)
SurvivedRF <- predict(rf.tune, newdata = pred.these)

Survived <- revalue(Survived, c("Survived" = 1, "Perished" = 0))
SurvivedRF <- revalue(SurvivedRF, c("Survived" = 1, "Perished" = 0))

predictions <- as.data.frame(Survived)
predictionsRF <- as.data.frame(SurvivedRF)

predictions$PassengerId <- df.infer$PassengerId
predictionsRF$PassengerId <- df.infer$PassengerId

write.csv(predictions[,c("PassengerId", "Survived")],
          file="Titanic_predictions.csv", row.names=FALSE, quote=FALSE)
write.csv(predictionsRF[,c("PassengerId", "SurvivedRF")],
          file="Titanic_predictionsRF.csv", row.names=FALSE, quote=FALSE)

createCSV <- function(tuning, data, fileName) {
  Survived <- predict(tuning, newdata = data)
  Survived <- revalue(Survived, c("Survived" = 1, "Perished" = 0))
  predictions <- as.data.frame(Survived)
  predictions$PassengerId <- df.infer$PassengerId
  write.csv(predictions[,c("PassengerId", "Survived")],
          file=fileName, row.names=FALSE, quote=FALSE)
}

createCSV(glm.tune.5, pred.these, "Titanic_predictions.csv")
createCSV(rf.tune, pred.these, "Titanic_predictionsRF.csv")
```
